Thesis Proposal

### 2019-09-25 ResearchReport.Rmd
Thesis proposal for feedback research seminar

-

>>>>

- Potential convergence measures:

*why do people say the GR stat will give false positives? perhaps the window (nr of it) is too small. if you start from dispersed, you'll end up with convergence (GR = 1) --> the bad thing is to stop too early. this is a false negative, not positive! possible approach: replicate and build a decision rule to solve the problem. *

*in mice algorithms, iterating is very fast compared to regular MCMC. therefore, we could use the average over the last 5 iterations you'll take the 'majority vote' to indicate conv. so compute GR at each iteration: band and threshold. another approach is to compute autocorrelation  (which is decreasing over itarations because we add in noise. schafer (1997, p. 129) wrote on worst linear stat and calculate the autocorr of that stat we know that the alg did converge. mice has noise so little autocorr. but we have very rich data so we're not estimating everything from scratch. not: we're talking about missing data only, not the combined data (that autocorr. is very high). autocorr is high if we can predict the miss value very well. the higher r sq the higher the autocorr. so the harder to converge. watch out for the problem with texp that will converge at once because you can deductively impute the value per class. possibly use slope of mean too to see whether there is trending. maybe also compute first eigenvalue of the matrix, not just mean and sd. also in pca we have several measures for nr of components, why not do something similar for mice? however, now we use mean and sd of missing values only, and that's not possible fr correlations because we'll have different ns for different correls. but of we take the compl. data that should not be a problem because that's constant. maybe look into allpying pca on the imputed data and if that stays the same we know that the means and variances are stable too. patrick ... and ian white work on stata proc mi also work on convergence diagnostics --> autocorr function plot in sas of worst linear function, and likelihood ratio that you could save.*

- Bayes course on convergence:
```{r}
# E. Assessing convergence
# -----------------------------------------------------------------------------
# History plot, autocorrelation plot
# History plots show the sampled parameters over the iterations (excluding
# the burn-in).
# The development/pattern in these plots gives an indication of convergence.
# When the history plot is stable (a fat catterpillar), convergence is reached.
# Autocorrelation can be measured at many lags.
# High autocorrelation indicates slow mixing of the random path.

# mcmcplot(mcmcout = samples)
# #sigma seems fine, but b doesn't:
# #even at lag 5 there is still quite some autocorrelation: therefore we need to center the predictors!
# #otherwise we could use a million iterations to deminish this effect

# Gelman-Rubin diagnostic
# The Gelman and Rubin statistic requires you run
# the sampler/algorithm at least twice: These runs are
# referred to as multiple chains.
# It compares the variance between chains to the variance
# within chains (G-R statistic = T/W = (pooled within chain
# var + between chain var)/pooled within chain var.
# It's the red line in the plot, and should be near 1.
# See Gibbs sampler presentation (week 2)
# https://drive.google.com/file/d/1ABHm8ala3c_puVvf32zF8h6TOZ3898uB/view
# pdf p. 41, slide nr 29.
# gelman.plot(samples)
# #this appears to be okay with a really small deviation from 1

# MC error (OPTIONAL?)
# MC error = SD/sqrt(number of iterations)
# SD represents the variation across iterations
# MC error thus represents how much the means differ w.r.t. the iterations
# MC error decreases as number of iterations increases.
# It should not be larger than 5% of the sample standard deviation

# History and density plot (OPTIONAL)
# plot(samples)

# Autocorrelation plots (OPTIONAL)
# autocorr.plot(samples)

# If parameters did not converge, you may:
# . Use (many) more iterations
# . Use a different parametrization (e.g., center predictors)
# . Use different priors (e.g., multivariate normal prior (i.e.,
#   dmnorm(,)) for parameters which are correlated)
# . Use other initial values
```

# Additional resources (not used)

**Potentially add something about updated (2019) version of $\widehat{R}$ and the new threshold of 1.01, see @veht19. Or leave it for the Research Report.**

"For models based on multiple imputed data sets, this [R hat > 1.1] is often a false positive: Chains of different submodels may not overlay each other exactly, since there were fitted to different data" (see https://cran.r-project.org/web/packages/brms/vignettes/brms_missings.html).

"Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chapter 11). It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" (@murr18, p. 19).

@gelm92's convergence criterion is defined as the ratio between the sum of the pooled within chain variance plus the between chain variance, and the pooled within chain variance.

"Why can the number of iterations in MICE be so low? First of all, realize that the imputed data Ymis form the only memory in the the MICE algorithm. Chapter 3 explained that imputed data can have a considerable amount of random noise, depending on the strength of the relations between the variables. Applications of MICE with lowly correlated data therefore inject a lot of noise into the system. Hence, the auto-correlation over t will be low, and convergence will be rapid, and in fact immediate if all variables are independent. Thus, the incorporation of noise into the imputed data has pleasant side-effect of speeding up convergence" (@buur18, par. 4.5).

### 2019-09-25 ThesisProposal.Rmd
Thesis proposal for feedback research seminar

---
title: "Thesis Proposal"
subtitle: "Methodology and Statistics for the Behavioural, Biomedical and Social Sciences"
author: "Hanne Oberman (4216318)"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    fig_caption: yes
bibliography: ThesisProposal.bib
---

@@

Word count:

@@

Requirements: max 750 words (excl. references)

In general the proposal should focus on the relevancy of the project, the gap in the scientific literature, and the feasibility to finish the project within 8 months. Possibly specify a 'step 2' to pursue after the main objective is reached.

@@

# Introduction/context and research question/goals

At some point, any (social) scientist conducting statistical analyses will run into a missing data problem (@alli02). Simply ignoring the missingness or using ad hoc solutions can yield wildly invalid inferences (@buur18). Multiple imputation (MI; @rubin87) provides a framework to circumvent the problem of missing information. This technique -- that is growing in popularity (@buur18) -- entails completing the missing values in an incomplete data set several times. The variability between the completed data sets captures the uncertainty in the inferences due to missingness (@rubin87).

The R package MICE (Multiple Imputation using Chained Equations; @mice11) is the world-leading software for multiple imputation (@buur18). However, the package currently does not provide user-friendly means to evaluate the multiply imputed data. Therefore, the goal of this project is develop a MICE evaluation suite, featuring existing modules (like plots to compare the incomplete and completed data sets), and novel assessment tools (like a measure to flag algorithmic non-convergence). The practical relevance of this research project is in facilitating applied researchers in making valid inferences despite of missingness. Simultaniously it contributes to the scientific literature because there is not yet a clear-cut way of diagnosing convergence for MI algorithms.

# Literature review

Currently, “there is no clear-cut method for determining when the MICE algorithm has converged” (@buur18, \textsection 6.5). Instead, users have to rely on visual inspection of the algorithm's iterations. Convergence is said to be reached when the parameters (like means of completed variables or regression coefficients) are stable over iterations (@whit11). 
<!-- "As MICE is an iterative procedure, it is important that convergence is achieved. This may be checked by computing, at each cycle, the means of imputed values and/or the values of regression coefficients, and seeing if they are stable" (@whit11, p. 394). -->
Additionally, convergence requires the variance between the several imputation chains to be no larger than the variances within each chain (@buur18). 
<!-- “Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence” (@buur18, \textsection 6.5).  -->
According to @li14 the latter is conceptually similar to the Gelman-Rubin statistic. 
<!-- "A common diagnostic tool is to plot one or more parameters against the iteration number and assess convergence by how different the variance between different sequences is relative to the variance within each individual sequence, similar to the Gelman-Rubin statistic (Gelman and Rubin, 1992) used in Markov chain Monte Carlo (MCMC) diagnostics" (@li14, \textsection 4.3). -->

The Gelman-Rubin statistic (@gelm92), however, assumes independence between chains. Therefore, it cannot be directly applied to compare imputations that share the same incomplete data (source??).

Notwithstandlingly, it seems like @su11 did implement the Gelman-Rubin statistic as convergence criterion into their R package 'mi', including the conventional cut-off of 1.1.
<!-- "R.hat: The value of the $\hat{R}$ statistic used as a convergence criterion. The default is 1.1 (Gelman and Rubin 1992; Gelman, Carlin, Stern, and Rubin 2004)" (@su11 p. 4). -->
<!-- "Our mi offers two ways to check the convergence of the multiple imputation procedure. By default, mi() monitors the mixing of each variable by the variance of its mean and standard deviation within and between different chains of the imputation. If the $\hat{R}$ statistic is smaller than 1.1, (i.e., the difference of the within and between variance is trivial), the imputation is considered converged (Gelman, Carlin, Stern, and Rubin 2004). Additionally, by specifying mi(data, check.coef.convergence = TRUE, ...), users can check the convergence of the parameters of the conditional models" (@su11, p. 13). -->
It is worth investigating its validity in the context of MICE's convergence, and the broader framework of fully conditional specification (FSC) algorithms.

The most recent literature on the topic states that convergence properties of FCS algorithms are still under debate (@taka17), with the specific case of a MICE procedure like predictive mean matching posing an entirely open question (@murr18).
<!-- "the convergence properties of FCS are currently under debate due to possible incompatibility (Li, Yu, and Rubin 2012; Zhu and Raghunathan 2015)" (@taka17, p. ) -->
<!-- "The convergence properties of FCS in general settings is still mostly an open question. The behavior of FCS algorithms under non- or quasi-Bayesian imputation procedures like PMM is entirely an open question" (@murr18, p.19) -->
@buur18 summarizes this problem as follows: "No method works best in all circumstances. The consensus is to assess convergence with a combination of tools. The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
<!-- “Several expository reviews are available that assess convergence diagnostics for MCMC methods (Cowles and Carlin 1996; Brooks and Gelman 1998; El Adlouni, Favre, and Bobée 2006). Cowles and Carlin (1996) conclude that “automated convergence monitoring (as by a machine) is unsafe and should be avoided.” No method works best in all circumstances. The consensus is to assess convergence with a combination of tools. The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5). -->
And that is exactly the gap in the scientific literature that this thesis will contribute to.

# Approach that will be used to answer the research question

During this research project I will work directly with developers of the MICE R package. I will use R Shiny (@shiny17) to program a web-browser based local app for the evaluation of multiply imputed data: 'ShinyMICE'. This approach does not require the use of any (personal) data. Therefore, the project will not be reviewed by the ethical committee.

@@

The research report that is to be handed in December 13th 2019 will consist of an investigation into algorithmic convergence of MICE. Ideally, this will result in a single indicator to flag non-convergence (potentially based on the Gelman-Rubin statistic, autocorrelation, or MC error).

The second stage of the research project is of more practical nature: programming ShinyMICE. The application will at least consist of the following: 1) one or more measures to assess algorithmic convergence; 2) data visualizations (e.g., scatterplots, densities, and cross-tabulations of the data pre- and post-imputation); and 3) statistical evaluation of relations between variables pre- versus post-imputation (i.e., chi square tests or t-tests). This part of the thesis also entails research into user interface (UI) design, local versus cloud-based data storage, and optimizing the application's efficiency.

@@

Finally, the shinyMICE package will be integrated into the existing MICE environment, and a vingette for applied researchers will be written.

# Additional resources (not used)

"Imputers who do choose to use FCS should use flexible univariate models wherever possible and take care to assess apparent convergence of the algorithm, for example by computing traces of pooled estimates or other statistics and using standard MCMC diagnostics (Gelman et al., 2013, Chpater [*sic*] 11). It may also be helpful to examine the results of many independent runs of the algorithm with different initializations and to use random scans over the p variables to try to identify any convergence issues and mitigate possible order dependence" (@murr18, p. 19).

<!-- @abay08 -->

<!-- @bart15 -->

<!-- @li91 -->

<!-- @rubin87 -->

<!-- @rubin96 -->

<!-- @vinknd -->

<!-- @scha02 -->

<!-- @cowl96 -->

<!-- @zhu15 -->

@gelm92's convergence criterion is defined as the ratio between the sum of the pooled within chain variance plus the between chain variance, and the pooled within chain variance.

**Proposal by supervisors:**

"The MICE package in R is a world-leading software package for multiple imputation. When using the mice function to solve missing data, vast amounts of information are calculated and stored. However, the MICE package currently lacks a user-friendly means of assessing this information. This project focuses on developing and programming novel means of evaluating, presenting and organizing this information in a web-browser based local app that allows for 1) comparing the imputed data to the observations, 2) inspecting the algorithmic convergence, 3) inspecting multivariate distributions, 4) the plausibility of the imputed data, and so on. During this project you will work closely with the developers of MICE in R and the coding components in this project will focus on R and Shiny.

The objective of the research is to 1) create a Shiny app to investigate and evaluate multiply imputed data sets, 2) implement it in MICE in R, 3) write an instructional vignette, 4) write a technical paper on the workings and usage of the software aimed at e.g. the R Journal or Journal of Statistical Software. The expected output is a state-of-the-art shiny app that can find its way into MICE".

**Bayes course on convergence:**
```{r}
# E. Assessing convergence
# -----------------------------------------------------------------------------
# History plot, autocorrelation plot
# History plots show the sampled parameters over the iterations (excluding
# the burn-in).
# The development/pattern in these plots gives an indication of convergence.
# When the history plot is stable (a fat catterpillar), convergence is reached.
# Autocorrelation can be measured at many lags.
# High autocorrelation indicates slow mixing of the random path.

# mcmcplot(mcmcout = samples)
# #sigma seems fine, but b doesn't:
# #even at lag 5 there is still quite some autocorrelation: therefore we need to center the predictors!
# #otherwise we could use a million iterations to deminish this effect

# Gelman-Rubin diagnostic
# The Gelman and Rubin statistic requires you run
# the sampler/algorithm at least twice: These runs are
# referred to as multiple chains.
# It compares the variance between chains to the variance
# within chains (G-R statistic = T/W = (pooled within chain
# var + between chain var)/pooled within chain var.
# It's the red line in the plot, and should be near 1.
# See Gibbs sampler presentation (week 2)
# https://drive.google.com/file/d/1ABHm8ala3c_puVvf32zF8h6TOZ3898uB/view
# pdf p. 41, slide nr 29.
# gelman.plot(samples)
# #this appears to be okay with a really small deviation from 1

# MC error (OPTIONAL?)
# MC error = SD/sqrt(number of iterations)
# SD represents the variation across iterations
# MC error thus represents how much the means differ w.r.t. the iterations
# MC error decreases as number of iterations increases.
# It should not be larger than 5% of the sample standard deviation

# History and density plot (OPTIONAL)
# plot(samples)

# Autocorrelation plots (OPTIONAL)
# autocorr.plot(samples)

# If parameters did not converge, you may:
# . Use (many) more iterations
# . Use a different parametrization (e.g., center predictors)
# . Use different priors (e.g., multivariate normal prior (i.e.,
#   dmnorm(,)) for parameters which are correlated)
# . Use other initial values
```

>>>>

---
title: "Thesis Proposal"
author: "Hanne Oberman (4216318)"
date: "`r Sys.Date()`"
output:
  #word_document: default
  pdf_document:
    fig_caption: yes
subtitle: Methodology and Statistics for the Behavioural, Biomedical and Social Sciences
bibliography: ThesisProposal.bib
---

@@

Word count: 821

@@

<!-- Requirements: max 750 words (excl. references) -->

<!-- In general the proposal should focus on the relevancy of the project, the gap in the scientific literature, and the feasibility to finish the project within 8 months. Possibly specify a 'step 2' to pursue after the main objective is reached. -->

@@

# Introduction

At some point, any (social) scientist conducting statistical analyses will run into a missing data problem [@alli02]. Simply ignoring the missingness or using ad hoc solutions can yield wildly invalid inferences [@buur18]. Multiple imputation (MI; [@rubin87]) provides a framework to circumvent the *ubiquitous* problem of missing information. This technique -- that is growing in popularity [@buur18] -- entails 'guessing' the missing values in an incomplete data set several times. The variability between the resulting completed data sets represents how much uncertainty in the inferences is due to missingness [@rubin87].

The R package MICE (Multiple Imputation using Chained Equations; [@mice11]) is the world-leading software for multiple imputation [@buur18]. However, the package currently does not provide user-friendly means to evaluate multiply imputed data. *This is vital because of the amount of assumptions in MICE algorithms. Thanks to technological developments in computing speed and data storage it has become possible to check assumptions where that was unfeasible before. The very first assumption to check is algorithmic convergence; without convergence, all 'deeper' assumptions and subsequent inferences are invalid.*

*The research question central to this project is: 'How can empirical researchers be facilitated in evaluating the  validity of multiply imputed data?'.* Therefore, the goal is to develop a MICE evaluation suite, featuring interactive adaptations of existing modules (like plots to compare the incomplete and completed data sets), and novel assessment tools (like a measure to flag algorithmic non-convergence). The practical relevance of this research project is in facilitating applied researchers in making valid inferences despite of missingness. Simultaneously, it contributes to the scientific literature because there is no definitive way of diagnosing non-convergence for MI algorithms.

# Literature Review

Since “there is no clear-cut method for determining when the MICE algorithm has converged” (@buur18, \textsection 6.5), users have to rely on visual inspection of the algorithm's iterations. Convergence is said to be reached when parameters (e.g., means of completed variables) are stable across iterations [@whit11]. 
<!-- "As MICE is an iterative procedure, it is important that convergence is achieved. This may be checked by computing, at each cycle, the means of imputed values and/or the values of regression coefficients, and seeing if they are stable" (@whit11, p. 394). -->
And additionally, the variation between imputation chains should be no larger than the variation within each individual chain [@buur18]. 
<!-- “Convergence is diagnosed when the variance between different sequences is no larger than the variance within each individual sequence” (@buur18, \textsection 6.5).  -->
*Conceptually, this visual evaluation procedure is similar to the 'Gelman-Rubin statistic' [@li14]. Practically, however, Gelman and Rubin's convergence diagnostic $\widehat{R}$ [-@gelm92] cannot be applied directly to MI data. The measure produces too many 'false alarms': chains that converged within ten iterations are still flagged as non-converged at iteration fifty [@lace07].*
<!-- "A common diagnostic tool is to plot one or more parameters against the iteration number and assess convergence by how different the variance between different sequences is relative to the variance within each individual sequence, similar to the Gelman-Rubin statistic (Gelman and Rubin, 1992) used in Markov chain Monte Carlo (MCMC) diagnostics" (@li14, \textsection 4.3). -->

Notwithstandingly, it seems like @su11 did implement $\widehat{R}$ as convergence criterion into their R package 'mi', including the conventional cut-off.
<!-- "R.hat: The value of the $\hat{R}$ statistic used as a convergence criterion. The default is 1.1 (Gelman and Rubin 1992; Gelman, Carlin, Stern, and Rubin 2004)" (@su11 p. 4). -->
<!-- "Our mi offers two ways to check the convergence of the multiple imputation procedure. By default, mi() monitors the mixing of each variable by the variance of its mean and standard deviation within and between different chains of the imputation. If the $\hat{R}$ statistic is smaller than 1.1, (i.e., the difference of the within and between variance is trivial), the imputation is considered converged (Gelman, Carlin, Stern, and Rubin 2004). Additionally, by specifying mi(data, check.coef.convergence = TRUE, ...), users can check the convergence of the parameters of the conditional models" (@su11, p. 13). -->
So it is worth investigating the validity of this convergence statistic in the context of multiple imputation.

The most recent literature on the topic states that convergence properties of MICE algorithms in general are still under debate [@taka17]--with specific procedures like 'predictive mean matching' posing entirely open questions [@murr18].
<!-- "the convergence properties of FCS are currently under debate due to possible incompatibility (Li, Yu, and Rubin 2012; Zhu and Raghunathan 2015)" (@taka17, p. ) -->
<!-- "The convergence properties of FCS in general settings is still mostly an open question. The behavior of FCS algorithms under non- or quasi-Bayesian imputation procedures like PMM is entirely an open question" (@murr18, p.19) -->
@buur18 summarizes this problem as follows: "No method works best in all circumstances. The consensus is to assess convergence with a combination of tools. The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5).
<!-- “Several expository reviews are available that assess convergence diagnostics for MCMC methods (Cowles and Carlin 1996; Brooks and Gelman 1998; El Adlouni, Favre, and Bobée 2006). Cowles and Carlin (1996) conclude that “automated convergence monitoring (as by a machine) is unsafe and should be avoided.” No method works best in all circumstances. The consensus is to assess convergence with a combination of tools. The added value of using a combination of convergence diagnostics for missing data imputation has not yet been systematically studied” (@buur18, \textsection 6.5). -->
And that is exactly the gap in the scientific literature that this thesis will contribute to.

# Methods

During this research project I will work directly with developers of the MICE R package. I will use R Shiny [@shiny17] to program an interactive evaluation device for the evaluation of multiply imputed data: 'ShinyMICE'. *This approach does not require the use of any empirical data. Therefore, the project does not need to be reviewed/approved by the faculty's ethical committee.*

@@

The research report that is to be handed in December 13th 2019 will consist of an investigation into algorithmic convergence of MICE. Ideally, this will result in a single indicator to flag non-convergence (potentially based on $\widehat{R}$, auto-correlation, or MC error).

The second stage of the research project is of more practical nature: programming ShinyMICE. The application will at least consist of the following: 1) one or more measures to assess algorithmic convergence; 2) data visualizations (e.g., scatter-plots, densities, and cross-tabulations of the data pre- and post-imputation); and 3) statistical evaluation of relations between variables pre- versus post-imputation (i.e., $\chi^2$-tests or t-tests). *This part of the thesis also entails research into user interface design, local versus cloud-based data storage, and optimizing the application's efficiency.*
<!-- 1) comparing the imputed data to the observations, 2) inspecting the algorithmic convergence, 3) inspecting multivariate distributions, 4) the plausibility of the imputed data -->

@@

Finally, the ShinyMICE package will be integrated into the existing MICE environment, and a vignette for applied researchers will be written.
